---
output: github_document
---

<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "man/figures/README-",
  out.width = "100%"
)
```

# stagetrackr

<!-- badges: start -->

<!-- badges: end -->

The goal of stagetrackr is to provide a range of tools to optimise working with data from species which develop through multiple life stages. It can automatically assign the last observed staged for individuals in experimental datasheets and plot graphs of: survival rates through development; distribution of last observed stages with the ability to specify factors which are separated for the purposes comparative analysis.

## Installation

You can install the development version of stagetrackr like so:

``` r
install.packages("devtools")
devtools::install_github("EcoHull/stagetrackr")
```

## Usage

The functions in this package are designed to be used in sequence depending on the questions being answered. But the base data relies on the recording of the dates that individuals move through developmental stages.

The first stage of stagetrackr analysis is to use the `assigned_stages()` function. This function takes the data frame and the stage order and returns the last observed developmental stage seen in each row, creating a column which is a factor with levels dictated by the provided columns.

```{r example}
library(stagetrackr)

example_stages = c("Stage1", "Stage2", "Stage3", "Stage4", "Stage5")

example_data = dplyr::tribble(
~ID, ~ Stage1, ~Stage2, ~Stage3, ~Stage4, ~Stage5,
  1, "01/01/2000", NA, NA, NA, NA,
  2, "01/01/2000", "02/01/2000", NA, NA, NA,
  3, "01/01/2000", "02/01/2000", "03/01/2000", NA, NA,
  4, "01/01/2000", "02/01/2000", "03/01/2000", "04/01/2000", NA,
  5, "01/01/2000", "02/01/2000", "03/01/2000", "04/01/2000", "05/01/2000",
  6, NA, NA, NA, NA, NA,
)

assigned_stages = stage_assigning(columns = example_stages, data = example_data)
show(assigned_stages)
```

Following the use of `stage_assigning()` a data table can be created using `last_stage_table()` this functions requires the specifying of the data frame, the column containing the last observed stages and a list of the stages in order and returns information about the stage distribution.

This function is designed to work with the output of `assigned_stages()` but will alwo work with any dataset with one row per individual and a column specifying final recorded developmental stage.

```{r stage_assigning_example}

data_table = last_stage_table(data = assigned_stages, last_observed_stage = "last_observed_stage")
show(data_table)
```

The `visualisig_survival()` function is designed to work with the output of `last_stage_table()`. It requires the specification of the data frame, remaining percentage column and optionally the remaining number column (if it is not provided it will not show the N value). This functions generates a bar chart with a bar for each developmental stage providing details of the remaining percentage (and optionally the N remaining) at each stage of development.

```{r visualising_survival_example}

visualising_survival(data = data_table, stages = last_observed_stage, remaining_percentage = remaining_percentage, remaining_number = remaining_n)

```

The distribution of the last observed stages can also be visualised. This used the `visualising_distribution()` function which is designed to work with the output of the `stage_assigning()` function. This requires the specification of the data frame, the column containing the assigned stages, and optionally a factor column. If a factor column is specified, a new bar will be generated for each factor.

```{r visualising_distrubution_no_factor_example, fig.height=6, fig.width=5}

# No factor visualisation

visualising_distribution(data = assigned_stages, stage = last_observed_stage)

```

```{r example_factorised_data}

# With factors specified
data_with_factor = dplyr::tribble(
  ~ID, ~ Stage1, ~Stage2, ~Stage3, ~Stage4, ~Stage5, ~last_observed_stage, ~Sex,
  1, "01/01/2000", NA, NA, NA, NA, "Stage1", "Male",
  2, "01/01/2000", "02/01/2000", "03/01/2000", NA, NA, "Stage3", "Male",
  3, "01/01/2000", "02/01/2000", "03/01/2000", NA, NA, "Stage3", "Male",
  4, "01/01/2000", "02/01/2000", "03/01/2000", "04/01/2000", NA, "Stage4", "Male",
  5, "01/01/2000", "02/01/2000", "03/01/2000", "04/01/2000", "05/01/2000", "Stage5", "Male",
  6, NA, NA, NA, NA, NA, "no_stage_found", "Male",
  1, "01/01/2000", NA, NA, NA, NA, "Stage1", "Female",
  2, "01/01/2000", "02/01/2000", NA, NA, NA, "Stage2", "Female",
  3, "01/01/2000", "02/01/2000", "03/01/2000", NA, NA, "Stage3", "Female",
  4, "01/01/2000", "02/01/2000", "03/01/2000", "04/01/2000", "05/01/2000", "Stage5", "Female",
  5, "01/01/2000", "02/01/2000", "03/01/2000", "04/01/2000", "05/01/2000", "Stage5", "Female",
  6, NA, NA, NA, NA, NA, "no_stage_found", "Female"
)

show(data_with_factor)
```

```{r visualising_distribution_example_with_factor}
visualising_distribution(data = data_with_factor, stage = last_observed_stage, factor = Sex)

```

### Time span analysis

For development time analysis first the data must be formatted. In order to do this it must first be handled as a date. This is done using `time_formatr` which requires a data frame, column and time format specification, this function uses the as.Date() function, this determines the specification required.

```{r time_formatr}

example_data = dplyr::tribble(
  ~ID, ~ Stage1, ~Stage2, ~Stage3, ~Stage4, ~Stage5, ~Sex,
  1, "01/01/2000", NA, NA, NA, NA, "Male",
  2, "01/01/2000", "03/01/2000", NA, NA, NA, "Male",
  3, "01/01/2000", "02/01/2000", "03/01/2000", NA, NA, "Male",
  4, "01/01/2000", "04/01/2000", "05/01/2000", "06/01/2000", NA, "Male",
  5, "01/01/2000", "02/01/2000", "04/01/2000", "07/01/2000", "10/01/2000", "Male",
  6, NA, NA, NA, NA, NA, "Female",
  7, "01/01/2000", NA, NA, NA, NA, "Female",
  8, "01/01/2000", "04/01/2000", NA, NA, NA, "Female",
  9, "01/01/2000", "03/01/2000", "05/01/2000", "09/01/2000", "12/01/2000", "Female",
  10, "01/01/2000", "05/01/2000", "06/01/2000", "07/01/2000", "11/01/2000", "Female"
)

formatted_time = time_formatr(example_data, example_stages, "%d/%m/%Y")
show(formatted_time)
```

Following formatting, time_spanr() can be used. This function takes the data frame and stages specified in a list and returns the time difference between each stage for each row.

```{r time_spanr}

time_span_data = time_spanr(formatted_time, example_stages)
show(time_span_data)

```

The result of this function is a new column for the time between each sequential developmental stage with a new column for each developmental period span. In order to visualise this distribution, `visualising_time_span()` can be used. This function takes the data frame, timespan columns and optionally a factor.

If a factor is not specified the development plot will show the distribution across all rows.

```{r time_spanr_example}

names = c("Stage1-Stage2", "Stage2-Stage3", "Stage3-Stage4", "Stage4-Stage5")

visualising_time_span(data = time_span_data, cols = names)

```

If a factor is specified a box will be shown for each variable fo the factor at each stage.

```{r time_spanr_factor}

visualising_time_span(data = time_span_data, cols = names, factor = Sex)

```
